---
title: "shrubling_vitals"
author: "Ruby An"
date: "2023-06-18"
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(googlesheets4)
library(knitr)

theme_set(theme_bw())
```

## Description
This code file plots data on focal shrubs planted in the shrub garden. 

```{r}
## Files 
treatment_key <- read_sheet("https://docs.google.com/spreadsheets/d/1PL6d-5xT28oVSgbeNmw1YzwA9chrVwz8SPmr5KjDAag/edit#gid=0", col_types = "dddcDll") %>% 
  select(Block, Plot, Block.Plot, Treatment) %>% 
  mutate(Snow = str_detect(Treatment, "S"), 
         OTC = str_detect(Treatment, "W"),
         Nutrients = str_detect(Treatment, "N"))

raw_data <- read_sheet("https://docs.google.com/spreadsheets/d/16xVSdXF8od_3aEwJrQ4FrA0SO1amAKwxkJEBk99hdVM/edit?usp=drive_link",
                       col_types = "dccddDDccc")

# Reshaping 
clean_data <- raw_data %>% 
  filter(!is.na(NewID)) %>% 
  left_join(treatment_key) %>% 
  mutate(Species = substr(NewID, 1,1)) 

## Shrub Origins
shrub_origins <- clean_data %>% select(Block.Plot, NewID, ShrubID) %>% 
  mutate(Origin = case_when(
    str_detect(str_c(ShrubID,NewID), "A") ~ "transplant",
    str_detect(str_c(ShrubID,NewID), "B") ~ "transplant",
    str_detect(ShrubID, "Sy\\d") ~ "transplant",
    str_detect(ShrubID, "Sy") ~ "connected", 
    str_detect(ShrubID, "^S$") ~ "trenched",
    TRUE ~ "unknown"
  ))

# unknown origin checks 
# shrub_origins %>% filter(Origin == "unknown") %>% print(n=30)

## Update clean_data with origin 
clean_data <- clean_data %>% 
  left_join(shrub_origins) %>% 
  mutate(Species.Origin = str_c(Species, ".", Origin))

```

## Experimental Set-up
This shows the set-up of the experiment and orients to the following graphs. 

```{r}
treatment_colors <- c("gray","chartreuse4","deepskyblue2", "aquamarine3","magenta3","purple3", "red2","darkorange")

ggplot(clean_data, aes(x=factor(Plot), y = factor(Block))) + 
  geom_tile(aes(fill = Treatment)) + 
    geom_text(aes(label = Block.Plot)) + 
  scale_fill_manual(values = treatment_colors) + 
  labs(x = "Plot", y = "Block", title = "Treatment Assignment") 


```


## Shrub Counts 

This code summarizes the number of shrubs found across all the plots. There is typically 1 alder, 2 birch, and 3 salix (transplant, trenched, natural) per plot. Some plots have 2 alders. Some plots have 3 birch. Some plots are missing a type of salix. 

### Struggle Shrubs

The following code plots the "struggle shrubs", noted in the inventory as "dead/questionable". Note that for retagging the shrubs, we chose the healthiest individuals, so these numbers will be an over-estimate of survival percentage. Nevertheless, we did try to tag at least 1 alder, 2 birch, and 3 kinds of salix in each plot. 

TREATMENT DEATH: There are more sick shrubs in snow removal and nutrient addition plots. There are fewer sick shrubs inside OTC's. 

PLOT DISTRIBUTION: All plots have 1-3 alders. All plots have 2-3 birch transplants. Salix are more spotty, with 11 plots are missing salix transplants. 6 plots are missing trenched salix. 5 plots are missing salix tagged as present. **note this will be updated when we check the suspicious salix inconsistencies. 

From this information, we can't clearly rule out using one or another type of Salix origin. For this, we must assess size and vitality. 

REPLANTING: I may need to replant 11 alder, 5 birch, and 9 salix. Only one S.connected and one S.trenched are dead, but I should check on the "retagging" situtation as I think we retagged quite a few S.trenched already so they are actually now S.connecteds. 

```{r}

## BONUS SHRUBS
#(YOU NEED TO REARRANGE THE SHEETS FOR THIS, WILL FIX SOON)
bonus_shrubs <- read_sheet("https://docs.google.com/spreadsheets/d/1PL6d-5xT28oVSgbeNmw1YzwA9chrVwz8SPmr5KjDAag/edit#gid=2120499410", sheet =2)

bonus_list <- bonus_shrubs %>% select(Block.Plot, Treatment, BonusShrubs) %>% 
  filter(!is.na(BonusShrubs)) %>% 
  separate_longer_delim(BonusShrubs, delim=";") %>% 
  mutate(BonusShrubs = str_trim(BonusShrubs)) %>% 
  mutate(Species = substr(BonusShrubs, 1,1)) %>% 
  arrange(Species, Block.Plot)

kable(bonus_list)
#write_csv(bonus_list, "bonus_shrub_list.csv")


# Sick Shrubs 
struggle_shrubs <- clean_data %>% filter(Vitality <= 2) %>%
  select(Block.Plot, NewID, Vitality, Symptoms, Treatment, Length_max, everything()) %>% 
  arrange(Block.Plot, Species,Vitality,Treatment) 

struggle_shrubs %>% 
  select(Block.Plot, Treatment, NewID, Vitality, Symptoms, Length_max,Notes) %>% kable()

#write_csv(struggle_shrubs, "struggle_shrub_list.csv")

# Sick List 

# sick_list <- shrub_inventory %>% filter(Vitality <=2) %>% select(Block.Plot, Treatment, NewID, Length_max, Vitality, Symptoms, Species) %>% arrange(Species, Treatment, Block.Plot, Vitality) 
# 
# #write_csv(sick_list, "sick_list_vit_less_2.csv")

# Sick Table
treatment_count <- clean_data %>% group_by(Species, Treatment) %>% 
  summarize(count = n())

shrub_count <- clean_data %>% group_by(Species.Origin) %>% 
  summarize(total_count = n())

struggle_shrubs %>% group_by(Species.Origin) %>% 
  summarize(sick_count = n()) %>% left_join(shrub_count) %>% 
  mutate(survival_percent = (total_count-sick_count)/total_count) %>%
  kable()
  
# Sick Plot Locations
struggle_shrubs %>% 
  select(Block, Plot, Species.Origin, NewID, Vitality) %>% 
  group_by(Block, Plot, Species.Origin) %>% 
  summarize(sick_count = n()) %>% 
  ggplot() + 
  geom_tile(aes(x=factor(Plot), y = factor(Block), fill= factor(Species.Origin))) + 
  labs(title = "Struggle Shrubs", x = "Plot", y = "Block") + 
  scale_fill_discrete(name = "Species")

# Sick Shrubs by Treatment
ggplot(struggle_shrubs) + 
  geom_histogram(aes(x = factor(Treatment), fill = Species), stat = "count") + 
  labs(x = "Block", y = "Sick Count", title = "Sick Shrubs by Treatment")

## Sick Shrubs by Block
ggplot(struggle_shrubs) + 
  geom_histogram(aes(x = factor(Block), fill = Species), stat = "count") + 
  labs(x = "Block", y = "Sick Count", title = "Sick Shrubs by Block")

## Sick Shrubs by Manipulation
ggplot(struggle_shrubs) + 
  geom_histogram(aes(x = Snow, fill = Species), stat = "count") + labs(x = "Snow Removal", y = "Sick Count", title = "Sick Shrubs by Snow Removal Treatment")

ggplot(struggle_shrubs) + 
  geom_histogram(aes(x = OTC, fill = Species), stat = "count") + labs(x = "Warming", y = "Sick Count", title = "Sick Shrubs by Warming Treatment")

ggplot(struggle_shrubs) + 
  geom_histogram(aes(x = Nutrients, fill = Species), stat = "count") + labs(x = "Nutrient Addition", y = "Sick Count", title = "Sick Shrubs by Nutrient Addition")

```

### Shrub Inventory
We can than assess the number of shrubs doing relatively well in each of the plots. The goal is to have at least 10 shrubs per treatment. For now, things are going relatively OK!  

```{r}
shrub_inventory <- clean_data %>% filter(!is.na(Vitality))

## TABLE BY TREATMENT
ggplot(shrub_inventory) + 
  geom_histogram(aes(x=factor(Block), fill = Vitality), stat = "count", position = "stack") + 
  geom_hline(yintercept = 8) + 
  facet_wrap(vars(Species)) + 
  labs(title = "Number of shrubs per Block")

## PLOT MAPS
shrub_inventory %>% 
  filter(!Species.Origin %in% c("S.unknown", "B.unknown")) %>% 
  group_by(Block, Plot, Species.Origin) %>% 
  summarize(count = n()) %>%
  ungroup() %>% 
  complete(Block, Plot, Species.Origin, fill = list(count = 0)) %>% 
  mutate(Block.Plot = str_c(Block, ".", Plot)) %>% 
  
  # viz
  ggplot(aes(x=factor(Plot), y = factor(Block))) + 
  geom_tile(aes(fill = factor(count))) + 
    geom_text(aes(label = Block.Plot)) + 
  scale_fill_manual(name = "Count", values = c("white", "skyblue", "blue", "darkblue")) + 
  facet_wrap(vars(Species.Origin)) + 
  labs(x = "Plot", y = "Block", title = "Number of shrubs per plot") 


## PLOT LOCATION OF VERY SMALL SHRUBS

## TABLES 
# ## count shrubs of each type in each plot
# plot_count <- clean_data %>% 
#   group_by(Block, Plot, Species, Origin) %>% 
#   summarize(count = n()) %>% 
#   pivot_wider(names_from = Plot, values_from = count) %>% 
#   select(Species, Origin, Block, everything())
# 
# ## alder count
# plot_count %>% filter(Species == "A") %>% kable()
# plot_count %>% filter(Species == "B") %>% kable()
# plot_count %>% filter(Species == "S" & Origin == "transplant") %>% kable()
# plot_count %>% filter(Species == "S" & Origin == "trenched") %>% kable()
# plot_count %>% filter(Species == "S" & Origin == "connected") %>% kable()
```


## Initial Shrub Sizes

We assessed the size of the shrubs to be able to standardize growth by initial biomass of each shrub. This will also help us know if we need to augment the experiment with any individuals by identifying short weaklings. 

On average, alders were 15 cm, birch were 10 cm, and salix were 13 cm. There was substantial variation in the salix. I grouped alders by size when planting, and tried to distribute evenly across treatments with similarly sized individuals within a block. This is reflected in the collected data. 

```{r}

## Table of means
mean_size <- clean_data %>% group_by(Species) %>% summarize(mean_size = mean(Length_max/10, na.rm = T), sd = sd(Length_max/10, na.rm= T))

kable(mean_size)

## overall size distribution 
ggplot(clean_data) + 
  geom_histogram(aes(y=Length_max/10, fill = Species), position = "identity") + 
  facet_wrap(vars(Species)) + 
  labs(x = "Count", y = "longest stem length (cm)", 
       title = "Shrubling Size Distribution")

# mean_size_by_block <- clean_data %>% group_by(Species, Block) %>% summarize(mean = mean(Length_max/10, na.rm = T), sd = sd(Length_max/10, na.rm= T))

# ## Plot barchart of means across blocks
# ggplot() +
#   ## bars
#   geom_bar(data = mean_size_by_block, aes(x=factor(Block), y = mean, fill = Species, alpha = 0.6), stat = "identity") + facet_wrap(vars(Species)) + 
#   
#   ## points 
#   geom_point(data = clean_data, aes(x = factor(Block), y = Length_max/10, color = Species)) + 
#   
#   labs(x = "Block", y = "longest stem length (cm)", title = "Shrub size distribution across blocks")
# 
# ## Plot of means across treatments
# 
# mean_size_by_treats <- clean_data %>% group_by(Species, Treatment) %>% summarize(mean = mean(Length_max/10, na.rm = T), sd = sd(Length_max/10, na.rm= T)) 
# ggplot() +
#   ## bars
#   geom_bar(data = mean_size_by_treats, aes(x=Treatment, y = mean, fill = Species, alpha = 0.6), stat = "identity") + facet_wrap(vars(Species)) + 
#   
#   ## points 
#   geom_point(data = clean_data, aes(x = Treatment, y = Length_max/10, color = Species)) + 
#   
#   labs(x = "Treatment", y = "longest stem length (cm)", title = "Shrub size distribution across treatments")


## BOXPLOTS of size distribuitions

## Plot boxplot of means across blocks
ggplot() +
  ## bars
  geom_boxplot(data = clean_data, aes(x=factor(Block), y = Length_max/10, fill = Species, alpha = 0.6)) + facet_wrap(vars(Species)) + 
  
  ## points 
  geom_point(data = clean_data, aes(x = factor(Block), y = Length_max/10, color = Species)) + 
  
  labs(x = "Block", y = "longest stem length (cm)", title = "Shrub size distribution across blocks")

## Boxplot of size across treatments 
ggplot() +
  ## bars
  geom_boxplot(data = clean_data, aes(x=Treatment, y = Length_max/10, fill = Species, alpha = 0.6)) + facet_wrap(vars(Species)) + 
  
  ## points 
  geom_point(data = clean_data, aes(x = Treatment, y = Length_max/10, color = Species)) + 
  
  labs(x = "Treatment", y = "longest stem length (cm)", title = "Shrub size distribution across treatments")
```

## Shrub Phenology 

This code plots the phenology for shrub transplants in 2023. We monitored for date of "buds breaking" and "leaves open". 
```{r}

phase_data <- clean_data %>% 
  pivot_longer(cols = "BB":"LO", 
               names_to = "Phase", 
               values_to = "Date") %>% 
  filter(!is.na(Date)) %>% 
  
  # count by date 
  ungroup() %>% 
  group_by(Treatment, Species, Phase, Date) %>% 
  summarize(phase_count = n()) %>% 
  ungroup() %>% 
  complete(Treatment, Species, Phase, Date, # complete missing dates
           fill = list(phase_count = 0)) %>% 
  
  # cumulative count by date
  group_by(Treatment, Species, Phase) %>% 
  mutate(phase_sum = cumsum(phase_count)) %>% 
  # % phase change 
  left_join(treatment_count) %>% 
  mutate(phase_percent = phase_sum/count) %>% 
  # treatment key code columns
  mutate(Snow = str_detect(Treatment, "S"), 
         OTC = str_detect(Treatment, "W"),
         Nutrients = str_detect(Treatment, "N"))


## All treatments overview
treatment_colors <- c("black","chartreuse4","deepskyblue2", "aquamarine3","magenta3","purple3", "red2","darkorange")
(phase_plot <- ggplot(phase_data, aes(x=Date, y = phase_percent, 
                                      color = Treatment, group = Treatment)) + 
   geom_point() + 
   geom_line() + 
    scale_color_manual(values = treatment_colors) + 
   facet_grid(Species ~ Phase))

## Snow removal impact on pheno
(snow_plot <- ggplot(phase_data, aes(x=Date, y = phase_percent, color = Snow, group = Treatment)) + 
  geom_point() + 
  geom_line() + 
    scale_color_manual(values = c("black", "deepskyblue2")) + 
    facet_grid(Species ~ Phase))

## Warming impact on pheno 
(otc_plot <- ggplot(phase_data, aes(x=Date, y = phase_percent, color = OTC, group = Treatment)) + 
    geom_point() + 
    geom_line() + 
    scale_color_manual(values = c("black", "red")) + 
    facet_grid(Species ~ Phase))

## Nutrient impact on pheno 
(nut_plot <- ggplot(phase_data, aes(x=Date, y = phase_percent, color = Nutrients, group = Treatment)) + 
    geom_point() + 
    geom_line() + 
    scale_color_manual(values = c("black", "chartreuse4")) + 

    facet_grid(Species ~ Phase))

```


## Shrub growth (biomass increase)

- photosynthetic canopy area
- new leaves, leaf size 

- height 
- new stems 
- height relative to competitors 

## Neighbors 

- species ID of neighbors 
- height and density of surrounding neighbors

## Environmental Neighborhood of Shrublings

- temperature, moisture
- light environment 
- nutrient availability 
- maximum snow depth 


